---
title: "ODEs and Nimble"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      cache = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      dpi = 600)
set.seed(0)
```

## Motivation

Here we show how to fit ODE-type models to noisy data using `Nimble`. We begin with writing a `C++` function encoding the ODE system and allowing for its numerical resolution. This step is done using the package `odeintr`. Then we define a `Nimble` function to wrap the `C++` ODE solver. We start with a 1D ODE example, then move to a 4D ODE for fun. Last, we fit a model that is specified with a system of ODEs and gaussian noise.


Let's load the packages:
```{r}
library(nimble)
library(odeintr)
library(Rcpp)
source("odetoNimble_fun.R")
```

## 1D example

Our first example is a simple logistic growth model $dy/dt= r y (1 - y/K)$. We define the system of equation then we build the `C++` code and compile it:
```{r}
logistic.sys = "dxdt[0] = alpha * x[0] * (1-x[0]/beta)"
myModel <- odetoNimble("logistic", logistic.sys, pars=c("alpha","beta"))
```

Now we use the `C++`model to define a new `Nimble`function:
```{r}
Cmodel_logistic <- compileNimble(myModel, showCompilerOutput = TRUE)
```

Set some values:
```{r}
y <- 0.1 # initial value
times <- seq(from = 0, to = 10, by = 0.01) # time sequence
params <- c(1.5, 10) # r and K
time_step = 0.01
```

Lets run the new function:
```{r}
resultat <- Cmodel_logistic(y, times, time_step,par=params)
head(resultat)
```

We now define a `Nimble` code:
```{r}
demoCode <- nimbleCode({
  init ~ dunif(0.1,0.1)
  y[1:1001,1:2] <- myModel(c(init),times[1:1001],0.01,param[1:2])
})
```

Build and compile le model:
```{r}
constants = list()
data = list()
inits = list(init=0.1,param=c(alpha=1.5,beta=10),times=times)
demoModel <- nimbleModel(demoCode, constants, data, inits, check = TRUE, calculate = FALSE)

CdemoModel <- compileNimble(demoModel, dirName=file.path(getwd(),"test"), showCompilerOutput = TRUE)

CdemoModel$calculate()
head(CdemoModel$y)
```

## 4D example

I use data from a paper by Witkowski and Brais entitled [Bayesian Analysis of Epidemics - Zombies, Influenza, and other Diseases](https://arxiv.org/abs/1311.6376). The authors provide a Python notebook [here](https://gist.github.com/bblais/181abd99f878282666b98a29588dda41). Briefly speaking, they counted the number of living deads in several famous zombie movies. I use the data from [Shaun of the Dead](https://www.imdb.com/title/tt0365748/).

The authors propose the following SIR epidemic model (more [here](https://github.com/oliviergimenez/SIRcovid19)) to capture the zombie apocalypse:
$$
dS/dt = - \beta S Z \\
dE/dt = \beta S Z - \zeta E \\
dZ/dt = \zeta E - \alpha S Z \\
dR/dt = \alpha S Z
$$
Let us solve this system of ODEs. First, set up some values:
```{r}
y <- c(508.2, 0, 1, 0) # intial values
times <- seq(from = 0, to = 50, by = 1) # time sequence
params <- c(0.2, 6, 0, 508.2) # beta, zeta, alpha, Sinit
```

Write the system of ODEs:
```{r}
zombie.sys = '
  dxdt[0] = -beta * x[0] * x[2] / Sinit;
  dxdt[1] = beta * x[0] * x[2] / Sinit - zeta * x[1];
  dxdt[2] = zeta * x[1] - alpha * x[0] * x[2];
  dxdt[3] = alpha * x[0] * x[2];
'
myModel <- odetoNimble("zombie", zombie.sys, pars=c("beta","zeta","alpha","Sinit"))
```

Write the `Nimble` code:
```{r}
code <- nimbleCode({
  xOde[1:51, 1:5] <- myModel(y[1:4], times[1:51],1, params[1:4])
})
constants <- list()
data <- list()
inits <- list(y = y, params = params, times=times)
demoModel <- nimbleModel(code, constants, data, inits, check = TRUE, calculate = FALSE)
CdemoModel <- compileNimble(demoModel, dirName=file.path(getwd(),"test"), showCompilerOutput = TRUE)
CdemoModel$calculate()
head(CdemoModel$xOde)
```

test avec `deSolve`:
```{r}
y <- c(508.2, 0, 1, 0) # intial values
times <- seq(from = 0, to = 50, by = 1) # time sequence
params <- c(0.2, 6, 0, 508.2) # beta, zeta, alpha, Sinit
shaun <- function(t, y, parms){
  dy1 <- - parms[1] * y[1] * y[3] / parms[4] # S
  dy2 <- parms[1] * y[1] * y[3] / parms[4] - parms[2] * y[2] # E
  dy3 <- parms[2] * y[2] - parms[3] * y[1] * y[3] # Z
  dy4 <- parms[3] * y[1] * y[3] # R
  return(list(c(dy1, dy2, dy3, dy4)))
} 
ode_nativeR <- deSolve::ode(y, times, shaun, params)[, -1]
head(ode_nativeR)
```

Compare with native `R`:
```{r}
sum(CdemoModel$xOde[,-1]-ode_nativeR)
```

It works, awesome ! 

## Add gaussian noise

Now let us have a look to a system with some observation error, which we assume is gaussian. 

First, we read in the data from *Shaun of the Dead*:
```{r}
tgrid <- c(0.00, 3.00, 5.00, 6.00, 8.00, 10.00, 22.00, 22.20, 22.50, 24.00, 25.50, 
           26.00, 26.50, 27.50, 27.75, 28.50, 29.00, 29.50, 31.50) 
zombies <- c(0, 1, 2, 2, 3, 3, 4, 6, 2, 3, 5, 12, 15, 25, 37, 25, 65, 80, 100)
```

Now the code:
```{r}
code <- nimbleCode({
  
  # system of ODEs
  xOde[1:ngrid, 1:(ndim+1)] <- myModel(y[1:ndim], times[1:ngrid], 0.1, params[1:ndim])
  # priors on parameters
  params[1] ~ dunif(0, 10) # beta
  params[2] ~ dunif(0, 1) # zeta
  params[3] ~ dunif(0, 0.01) # alpha
  params[4] ~ dunif(300, 600) # Sinit
  
  # observation error
  for (i in 1:ngrid){
    obs_x[i] ~ dnorm(xOde[i, 4], tau.x)
  }
  
  # prior on error sd
  tau.x <- 1 / var.x
  var.x <- 1 / (sd.x * sd.x)
  sd.x ~ dunif(0, 5)
})
```

Specify the constants, data and initial values:
```{r}
# constants
constants <- list(ngrid = 19, 
                  ndim = 4)

# data (pass times and y in constants?)
data <- list(times = tgrid,
             obs_x = zombies,
             y = c(508.2, 0, 1, 0))

# initial values
inits <- list(sdx = 2)
```

Get ready:
```{r}
Rmodel <- nimbleModel(code, constants, data, inits)
# Rmodel$calculate()   ## NA...
conf <- configureMCMC(Rmodel)
conf$printMonitors()
```

```{r}
conf$printSamplers(byType = TRUE)
```

```{r}
Rmcmc <- buildMCMC(conf)
Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
```

Unleash the beast:
```{r}
samplesList <- runMCMC(Cmcmc, 5000, 
                       nburnin = 1000,
                       thin = 10,
                       nchains = 2,
                       samplesAsCodaMCMC = TRUE)
```

Check out convergence:
```{r}
library(coda)
gelman.diag(samplesList)
```

Visualize traceplots and posterior distributions:
```{r}
library(basicMCMCplots)
chainsPlot(samplesList)
```

Apart from the standard deviation of the observation error, the mixing is poor. This is what I had with `OpenBUGS` too, see [here](https://github.com/oliviergimenez/fitODEswithOpenBUGS/blob/master/README.md). 

Can we do something about that? Hopefully yes, and this is what's great with `Nimble`, you have full control of the underlying MCMC machinery. There are useful advices on the Nimble forum [here](https://groups.google.com/forum/#!searchin/nimble-users/improve$20mixing%7Csort:date/nimble-users/pggH93kjBeE/OCj7STZWAgAJ). One reason for poor mixing is correlation in parameters. Let's have a look then. 
```{r}
cor(samplesList$chain1)
cor(samplesList$chain2)
```

Seems like some parameters are slightly correlated, like $\beta$ (params[1]) and $\zeta$ (params[2]) for example. For the sake of example, let's use block sampling to try and improve mixing. 

First, ask what are the samples used currently: 
```{r}
conf$printSamplers()
```

OK, now remove the default samplers for params[1] and params[3] and use a block random walk instead:
```{r}
conf$removeSamplers(c('params[1]', 'params[2]'))
conf$printSamplers()
conf$addSampler(target = c('params[1]', 'params[2]'),
                type = 'RW_block')
conf$printSamplers()
```

Now rebuild, recompile:
```{r}
Rmcmc <- buildMCMC(conf)
Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
```

Now rerun:
```{r}
samplesList <- runMCMC(Cmcmc, 5000, 
                       nburnin = 1000,
                       thin = 10,
                       nchains = 2,
                       samplesAsCodaMCMC = TRUE)
```

Visualize traceplots and posterior distributions:
```{r}
chainsPlot(samplesList)
```

What if we use another sampler?
```{r}
Rmodel <- nimbleModel(code, constants, data, inits)
conf <- configureMCMC(Rmodel)
conf$printSamplers()
conf <- configureMCMC(Rmodel, onlySlice = TRUE)
conf$printSamplers()
```

Now rebuild, recompile:
```{r}
Rmcmc <- buildMCMC(conf)
Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
```

Now rerun:
```{r}
samplesList <- runMCMC(Cmcmc, 2000, 
                       nburnin = 1000,
                       thin = 1,
                       nchains = 2,
                       samplesAsCodaMCMC = TRUE)
```

Visualize traceplots and posterior distributions:
```{r}
chainsPlot(samplesList)
```

Much slower.

## R version used

```{r}
sessionInfo()
```

